---
title: "Fish List"
author: "Derrick Theophille"
date: "4 April 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a short exercise that follows the stages of building an "all-purpose" fish species dataset for use in your further fisheries analysis.

I will do my best to explain every step of the journey with you, so look out for the hashtags (#) within the code chunks (the sections that begin with ```{}), they offer explanations for what is going on.

## Required Libraries
Before we begin, we need to load a few libraries that will help make the work easier. Libraries are add-ons that offer more functionality to the basic R. In the chunk below I 

```{r required libraries}
# Install the packages I will be using. I plan to use "dplyr", "tidyr" and "knitr". You may have already installed these packages. But in case you haven't you can do so now by running the line below.
# install.packages(c("dplyr", "tidyr", "knitr", "repmis"))

# Load the installed packages into our R session. Explanations given after hashtags. 
library(dplyr) # for summaries
library(tidyr) # for reshaping data
library(knitr) # for neat tables
library(repmis) # for accessing data from online. Keep an eye on the prompt in the console window (if you use RStudio, it is the bottom-left window) where you will be asked to select "Y" or "n", after reading the message. In order to proceed, however, you will need to type "Y".
```

## Required Datasets

```{r my fish list}
# My local fish list is stored in a binary file called "LUfish.rda". For your convenience I've made it available only from my Git repository. Run the code below to download and attach it.
source_data("https://github.com/theode/UFD_Interactions/blob/master/LUfish.rda?raw=true")

# Let's look at the data briefly to get an idea what is included:
head(LUfish)

# The column names aren't the most intuitive for humans, but that's because the data was ripped directly from a database query, where the names were not meant for common consumption, only the database manager. We can fix this up:
LUfish <- rename(LUfish, # We are going to edit some of the column names of LUfish
                 ID = fishID, # ID replaces fishID
                 English = fishCommName, # English replaces fishCommName
                 Fishery = fishFisheryCodeLU, # Fishery replaces fishFisheryCodeLU
                 Patois = fishPatName, # Patois replaces fishPatName
                 Scientific = fishSciName) # Scientific replaces fishSciName

# The "Group" and "Category" column labels cna remain. These are used for sorting and grouping fish species in some of my other reports.

# Lets look at a summary of the data
summary(LUfish)
# This isn't very helpful as our data does not have any useful numerical variables that can give a summary. This does, however show some useful information. Take a look and see for yourself.

glimpse(LUfish)
```

```{r FAO ASFIS data}
# I converted the xlsx file Mario shared on the Dgroup to an rda format and stored on Github. Access and load here:
source_data("https://github.com/theode/UFD_Interactions/blob/master/asfis.rda?raw=true")

head(asfis)

glimpse(asfis) # There are almost 13,000 observations in this dataset and 14 variables (columns). I Do not think we'll be needing Russian and Chinese names any time soon, so we will remove them from our dataset. We'll just keep the 3 Alpha code, the scientific name, the english name, the author, family and order.

asfis <- select(asfis,
                `3A_CODE`, Scientific_name, English_name, Author, Family, Order)
# Note that the variable 3A_CODE needs to be enclosed in backquotes (`). This is because 3A_CODE is a non-standard variable name as it begins with a number. Backquotes are used to allow the use of variables such as this.
```

```{r join the datasets}
# Now for the fun part: joining the data. 
# I think it's best to join the data using the scientific name variale as this is most-likely the common variable in both datasets. The only problem is that the variable is called slightly differnet names in both datasets. No problem, though, we can tell R which variables to use within the "by=c()" argument.
fishlist <- LUfish %>% 
  left_join(asfis, by=c("Scientific" = "Scientific_name"))

str(fishlist)

# Note also that the final data is only 166 observations (or species) and 12 variables. Why do you think that is?

# Lets make a more streamlined table in preparation of our final task.
fishlist <- select(fishlist,
                   `3A_CODE`, English_name, Patois, Scientific, Author) %>% 
  arrange(English_name)
```

```{r table}
# Finally, lets prepare a fancy table for our list.
kable(fishlist)
```

